Parallel global optimization algorithm with uniform convergence for solving a set of constrained global optimization problems\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-06-13 2020/OPTIMA/optima_paper.tex
Check Section 1 and 2

\begin{document}
%
\title{Parallel global optimization algorithm for obtaining uniform convergence when
simultaneously solving a set of global optimization problems\thanks{This study was supported
by the Russian Science Foundation, project No.\,16-11-10150.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Vladislav Sovrasov\and
Konstantin Barkalov
}
%
\authorrunning{V. Sovrasov et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia
\email{sovrasov.vladislav@itmm.unn.ru},
\email{konstantin.barkalov@itmm.unn.ru}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In the present work, an extension of global optimization algorithm solving a number of problems
simultaneously onto a case of problems with non-linear constraints is considered. The property
of uniform convergence inherent to the algorithm allows distributing the computation resources
in the optimal way since the errors of numerical solutions decrease with approximately equal
rates in all problems in the course of the execution of the algorithm. The algorithm assigns a
priority to each problem and performs the computations of the objective functions in several
problems in parallel at every iteration. Upon stop of the method at arbitrary time moment,
solutions of similar quality would be obtained in all problems of the series being solved. A series
of several problems arises if the global optimization problem has a discrete parameter or, for
example, in solving a multicriteria optimization problem using criteria convolution. The
considered algorithm employs mappings like Peano space filling curve for the reduction of the
multidimensional optimization problems to the one-dimensional ones.  The efficiency of the
implemented algorithm was tested using the sets of artificially generated global optimization
problems as well as in solving a series of problems obtained as a result of scalarization of a
multicriteria optimization  problem. Also, the numerical experiments confirmed the uniform
convergence of the method.

\keywords{Global optimization \and Uniform convergence \and Derivative-free optimization
\and Parallel computations.}
\end{abstract}
%
%
%

\section{Introduction}
\label{sec-intro}
Nonlinear global optimization of the non-convex functions is considered traditionally to be one
of the most difficult problems of mathematical programming.
 The finding of the global minimum of a function of several variables appears to be more
difficult often than a local optimization in a thousand-dimensional space.
For the latter, application of the simplest gradient descent method may appear to be enough
whereas in order to \textit{guarantee} the finding of the global optimum, the optimization
methods have to accumulate the information on the behavior of the objective function in the
whole search domain \cite{Jones2009,Paulavicius2011,Evtushenko2013,Strongin2000}.
The solving of a series of such problems at limited computational resources is even more
difficult problem: besides the search of the global extremum, it is necessary to distribute the
computational resources in such a way that the location of the global extremum would be
estimated with approximately equal quality in all problems being solved at once.
Usually, a series of \(q\) problems is solved either sequentially or in parallel by portions of \(p\ll
q\) problems where \(p\) is a number of parallel computational devices.
Such an approach leads to a situation when at every time moment until the stop of computations
some problems remain, in which the estimate of the global optimum is not obtained yet whereas
in the problems from the beginning of the list the optimum may be estimated even with excess
precision.

In the present work, a generalization of parallel global optimization method for simultaneous
solving of a set of problems developed earlier in Lobachevskii University of Nizhnii Novgorod
\cite{BarkalovStrongin2018} onto the case of nonlinear constraints is considered. To account
for the constraints, the index scheme \cite{Strongin2000} is applied allowing working with
partially computable objective functions and featured by efficiency comparable to other
approaches \cite{BarkalovLebedev2017}. The convergence of the algorithm to the global
optimal solutions in all problems was proved theoretically, the efficiency of the implemented
algorithm was demonstrated by solving sets of problems generated by a special mechanism
generating the sets of problems of predefined dimensionality with predefined number of
nonlinear constraints \cite{GergelBarkalov2019}.
Besides the problems generated artificially, the considered method was tested also on a set of
problems arising when solving a multicriteria optimization problem with nonlinear constraints by
criteria convolution method \cite{Ehrgott2005}.

@@

\section{Global optimization problem statement}
\label{sec:problem}
Within the frames of the present work, we will consider the following global optimization
problem statement: to find the global minimum of an \(N\)-dimensional function \(\varphi(y)\) in
a hyperinterval
\(D=\{y\in \mathbf{R}^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\).
In order to construct an estimate of the global minimum based on a finite number of
computations of the function values, the rate of variation of \(\varphi(y)\) in \(D\) is required to
be limited.
As a rule, the Lipschitz condition Is accepted as such a limitation.
\begin{equation}
\label{eq:task}
\varphi(y^*)=\min\{\varphi(y):y\in D\}
\end{equation}
\begin{displaymath}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty
\end{displaymath}

There are various methods solving the considered multidimensional problem directly
\cite{SergeyevKvasov2017, Jones2009} as well as efficient methods of solving the one-
dimensional problems \cite{Norkin1992, Strongin2000}.
In the present work, a one-dimensional method is considered, which is applied together with the
dimensionality reduction scheme.
The use of evolvents or space-filling curves is a classical scheme of reduction of the
dimensionality of the initial problem for the global optimization algorithms
\cite{Sergeyev2013}.
\begin{equation}
\label{cube}
\lbrace y\in \mathbf{R}^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant
N\rbrace=\{y(x):0\leqslant x\leqslant 1\}
\end{equation}

A mapping of the kind (\ref{cube}) allows reducing a problem in the multidimensional space to
solving a one-dimensional one at the expense of worsening its properties. In particular, the one-
dimensional function \(\varphi(y(x))\) is not a Lipschitz but a H\"{o}lder one:
\begin{displaymath}
\label{holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0;1],
\end{displaymath}
where the H\"{o}lder constant \(H\) is related to the Lipschitz one \(L\) by the relation
\begin{displaymath}
  H=2L\sqrt{N+3}.
\end{displaymath}

The domain \(D\) can also be defined by the constraint functions that complicates the problem
essentially.
The statement of the global optimization problem in this case will take the following form:
\begin{equation}
  \label{eq:constrained_problem}
  \varphi(y^*)=\min\{\varphi(y):y\in G=D \cap \{y: g_j(y)\leqslant 0, 1\leqslant j\leqslant m\}\}
\end{equation}

Let us denote \(g_{m+1}(y)=\varphi(y)\). Hereafter, we will assume all functions
\(g_k(y),1\leqslant k \leqslant m+1\)
to satisfy the Lipschitz condition in some hyperinterval including \(D\).

@@

Taking into account the dimensionality reduction scheme (\ref{cube}), let us assume when
describing the method that it is required to find the global minimum of the function \(\varphi(x),
x\in[0;1]\) satisfying the H\"{o}lder condition with the constraints \(g_j(x)\) also satisfying this
condition in the interval \([0;1]\).

The index algorithm of global search (IAGS) for solving the one-dimensional problems
(\ref{eq:constrained_problem}) considered here implies the constructing of a sequence of points
\(x_k\), which the values of the minimized function or of the constraints \(z_k = g_s(x_k)\) are
computed in.
To account for the latter, the index scheme \cite{Strongin2000} was used.
Let \(Q_0=[0;1]\). The constraint having the index \(j\) is satisfied in all points of the domain
\begin{displaymath}
  Q_j=\left\{x\in [0;1]:g_j(x)\leq 0\right\},
\end{displaymath}
which is called feasible for this constraint.
At that, the feasible domain \(D\) of the initial problem is defined by the equality:
 \(D=\cap _{j=0}^{m}Q_{j}\).
A trial in the point \(x\in [0;1]\) consists in a sequence computing of the values
\(g_{1}(x),...,g_{\nu }(x)\)
where the values of the indices \(\nu\) are defined by the conditions:
\(x\in Q_{j},0\leqslant j<\nu ,x\notin Q_{\nu }\).
The finding of the first violation of any constraint terminates the trial in the point \(x\).
In the case when the point \(x\) is feasible i. e. \(x\in D\), the trial includes the computing of all
functions of the problem.
At that, the value of the index is accepted to be equal to the magnitude \(\nu =m+1\).
The pair \(\nu =\nu (x), z=g_{\nu }(x)\) where the index \(\nu\) falls into the borders \(1\leqslant
\nu \leqslant m+1\) is called the result of the trial in the point \(x\).

@@

\begin{displaymath}
  \begin{array}{lr}
    \psi (x^{*})=\min_{x\in [0;1]}\psi (x), \\
    \psi (x)={\begin{cases}g_{\nu }(x)/H_{\nu }&\nu <M\\(g_{M}(x)-g_{M}^{*})/H_{M}&\nu
=M\end{cases}}
  \end{array}
\end{displaymath}

Here \(M=\max_{}^{}\left\{\nu (x):x\in [0;1]\right\}\) and \(g_{M}^{*}=\min
_{}^{}\left\{g_{M}(x):x\in \cap _{i=0}^{M-1}Q_{i}\right\}\).
Because of the definition of the number \(M\), the problem of finding \(g_{M}^{*}\) always
has a solution. If \(M=m+1\), then \(g_{M}^{*}=\varphi(x^{*})\).
The arcs of the functions \(\psi (x)\) are the H\"{o}lder ones within the sets
\(\cap _{i=0}^{j}Q_{i},0\leq j\leq M-1\)
with the constant 1, and \(\psi (x)\) itself can have the breaks of the first kind at the boundaries
of these sets.
Despite the values of the H\"{o}lder constants \(H_k\) and the value \(g_{M}^{*}\) are not
known in advance, these ones can be estimated in the course of solving the problem.

@@

At the first iteration of the method, the trial is performed in the arbitrary internal point \(x_1\) of
the interval \([0;1]\).
The indices of the points 0 and 1 are considered to be zero ones, the values \(z\) in these points
are undefined.
Let \(k\geqslant 1\) iterations of the method have been performed, in the course of which the
trials were conducted in \(k\) points \(x_i, 1\leqslant i\leqslant k\).
Then, the points \(x^{k+1}\) of the search trials of the next \((k+1)^{\rm th}\) iteration are
defined according to the rules:

Step 1. Renumber the points of the set \(X_k=\{x^1,\dotsc,x^k\}\cup\{0\}\cup\{1\}\), which
includes the boundary points of the interval \([0;1]\) as well as the points of preceding trials by
the lower indices in the order of increasing coordinate values i. e.
\begin{equation}
  \label{eq:points}
  0=x_0<x_1<\dotsc<x_{k+1}=1
\end{equation}
and juxtapose the values \(z_{i}=g_{\nu }(x_{i}),\nu =\nu (x_{i}),i={\overline {1,k}}\) to these
ones.

@@

Step 7. Perform the next trial in the middle of the interval \((x_{t-1};x_{t})\) if the indices of its
end points are not the same: \(x^{k+1}={\frac {1}{2}}(x_{t}+x_{t-1})\).
Otherwise, perform the trial in the point
\begin{displaymath}
  x^{k+1}={\frac {1}{2}}(x_{t}+x_{t-1})-\operatorname {sgn}(z_{t}-z_{t-1}){\frac {|z_{t}-
z_{t-1}|^{n}}{2r_{\nu }\mu _{\nu }^{n}}},\nu =\nu (x_{t})=\nu (x_{t-1}),
\end{displaymath}
and then increment \(k\) by 1.

>>>>

\begin{document}
%
\title{Parallel global optimization algorithm with uniform convergence
for solving a set of constrained global optimization problems\thanks{This study was supported
by the Russian Science Foundation, project No.\,16-11-10150.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Vladislav Sovrasov\and
Konstantin Barkalov
}
%
\authorrunning{V. Sovrasov et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia
\email{sovrasov.vladislav@itmm.unn.ru},
\email{konstantin.barkalov@itmm.unn.ru}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In the present work, an extension of global optimization algorithm for solving a set of problems
simultaneously onto a case of problems with non-convex constraints is considered. The property
of uniform convergence inherent to the algorithm allows distributing the computation resources
in the optimal way since the errors of numerical solutions decrease with approximately equal
rates in all problems in the course of the execution of the algorithm. The algorithm assigns a
priority to each problem and performs the computations of the objective functions in several
problems in parallel at every iteration. Upon stop of the method at arbitrary time moment,
solutions of similar quality would be obtained in all problems of the series being solved. A series
of several problems arises if the global optimization problem has a discrete parameter or, for
example, in solving a multicriteria optimization problem using criteria convolution. The
considered algorithm employs Peano-type space-filling curves for the reduction of the
multidimensional optimization problems to the one-dimensional ones.  The efficiency of the
implemented algorithm was tested using the sets of artificially generated global optimization
problems as well as in solving a series of problems obtained as a result of scalarization of a
multicriteria optimization  problem. Also, the numerical experiments confirmed the uniform
convergence of the proposed method.

\keywords{Global optimization \and Non-convex constraints \and Uniform convergence \and Derivative-free optimization
\and Parallel computing.}
\end{abstract}
%
%
%

\section{Introduction}
\label{sec-intro}
Global optimization problems with non-convex constraints are considered traditionally to be one
of the most difficult optimization problems.
 The finding of the global minimum of a function of several variables appears to be more
difficult often than a local optimization in a thousand-dimensional space.
For the latter, application of the simplest gradient descent method may appear to be enough,
whereas to \textit{guarantee} the finding of the global optimum the optimization
methods have to accumulate the information on the behaviour of the objective function in the
whole search domain \cite{Jones2009,Paulavicius2011,Evtushenko2013,Strongin2000}.
The solving of a series of such problems at limited computational resources is even more
difficult problem: besides the search of the global extremum, it is necessary to distribute the
computational resources in such a way that the location of the global extremum would be
estimated with approximately equal quality in all problems being solved at once.
Usually, a series of \(q\) problems is solved either sequentially or in parallel by portions of \(p\ll
q\) problems where \(p\) is a number of parallel computational devices.
Such an approach leads to a situation when at every time moment until the stop of computations
some problems remain, in which the estimate of the global optimum is not obtained yet whereas
in the problems from the beginning of the list the optimum may be estimated even with excess
precision.

In the present work, a generalization of parallel global optimization method for simultaneous
solving of a set of problems developed earlier in Lobachevsky State University of Nizhni Novgorod
\cite{BarkalovStrongin2018} onto the case of non-convex constraints is considered. To account
for the constraints, the index scheme \cite{Strongin2000,Pugliese} is applied allowing working with
partially defined objective function and constraints and featured by efficiency comparable to other
approaches \cite{BarkalovLebedev2017}. The convergence of the algorithm to the global
optimizers in all problems was proved theoretically, the efficiency of the implemented
algorithm was demonstrated by solving sets of problems generated by a special mechanism
generating the sets of problems of predefined dimensionality with predefined number of
non-convex constraints \cite{GergelBarkalov2019}.
Besides the problems generated artificially, the considered method was tested also on a set of
problems arising when solving a multicriteria optimization problem with nonlinear constraints by
criteria convolution method \cite{Ehrgott2005}.

@@

\section{Global optimization problem statement}
\label{sec:problem}
In this work we consider the following 
problem: to find the global minimum of an \(N\)-dimensional function \(\varphi(y)\) in
a hyperinterval $D$
\begin{equation}
\label{eq:task}
\varphi(y^*)=\min\{\varphi(y):y\in D\},
\end{equation}
\begin{displaymath}
D=\{y\in \mathbf{R}^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}.
\end{displaymath}
In order to construct an estimate of the global minimum based on a finite number of
computations of the function values, the rate of variation of \(\varphi(y)\) in \(D\) is required to
be limited.
As a rule, the Lipschitz condition 
\begin{displaymath}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty,
\end{displaymath}
is accepted as such a limitation.

There are various methods solving the considered multidimensional problem directly
\cite{SergeyevKvasov2017, Jones2009} as well as efficient methods for solving the univariate problems \cite{Norkin1992, Strongin2000}.
In the present work, a one-dimensional method is considered, which is applied together with the
dimensionality reduction scheme.
The use of space-filling curve (or \textit{evolvent}) $y(x)$, where
\begin{equation}
\label{cube}
\lbrace y\in \mathbf{R}^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant
N\rbrace=\{y(x):0\leqslant x\leqslant 1\},
\end{equation}
 is a classical scheme of reduction of the
dimensionality of the initial problem for the global optimization algorithms
\cite{Sergeyev2013}.
A mapping of the type (\ref{cube}) allows reducing a multidimensional problem to a univariate one at the expense of worsening its properties. In particular, the 
one-dimensional function \(\varphi(y(x))\) is not lipschitzian but h\"{o}lderian one:
\begin{displaymath}
\label{holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}}, \; x_1,x_2\in[0,1],
\end{displaymath}
where the H\"{o}lder constant \(H\) is related to the Lipschitz one \(L\) by the relation
\begin{displaymath}
  H=2L\sqrt{N+3}.
\end{displaymath}

The feasible domain \(G\) can also be defined by the constraints that essentially complicate the problem.
The problem statement in this case will take the following form:
\begin{equation}
  \label{eq:constrained_problem}
  \varphi(y^*)=\min\{\varphi(y):y\in G=D \cap \{y: g_j(y)\leqslant 0, 1\leqslant j\leqslant m\}\}.
\end{equation}

Let us denote \(g_{m+1}(y)=\varphi(y)\). Hereafter, we will assume all functions
\(g_k(y),1\leqslant k \leqslant m+1\),
to satisfy the Lipschitz condition in some hyperinterval including \(D\).

@@

Taking into account the dimensionality reduction scheme (\ref{cube}), let us assume when
describing the method that it is required to find the global minimum of the function \(\varphi(x),
x\in[0,1]\), satisfying the H\"{o}lder condition with the constraints \(g_j(x)\) also satisfying this
condition in the interval \([0,1]\).

The index algorithm of global search (IAGS) for solving the one-dimensional problems
%(\ref{eq:constrained_problem}) 
considered here implies the constructing of a sequence of points
\(x_k\), which the values of the minimized function or of the constraints \(z_k = g_s(x_k)\) are
computed in.
To account for the latter, the index scheme \cite{Strongin2000} was used.
Let \(Q_0=[0,1]\). The constraint having the index \(j\) is satisfied in all points of the domain
\begin{displaymath}
  Q_j=\left\{x\in [0,1]:g_j(x)\leq 0\right\},
\end{displaymath}
which is called feasible for this constraint.
At that, the feasible domain \(D\) of the initial problem is defined by the equality
 \(D=\cap _{j=0}^{m}Q_{j}\).
A trial in the point \(x\in [0,1]\) consists in a sequence computing of the values
\(g_{1}(x),...,g_{\nu }(x)\)
where the values of the indices \(\nu\) are defined by the conditions:
\(x\in Q_{j},0\leqslant j<\nu ,x\notin Q_{\nu }\).
The finding of the first violation of any constraint terminates the trial at the point \(x\).
In the case when the point \(x\) is feasible, i.e. \(x\in D\), the trial includes the computing of all
functions of the problem.
At that, the value of the index is accepted to be equal to the magnitude \(\nu =m+1\).
The pair \(\nu =\nu (x), z=g_{\nu }(x)\) where the index \(\nu\) falls into the borders \(1\leqslant
\nu \leqslant m+1\) is called the result of the trial at the point \(x\).

@@

\begin{displaymath}
  \begin{array}{lr}
    \psi (x^{*})=\min_{x\in [0,1]}\psi (x), \\
    \psi (x)={\begin{cases}g_{\nu }(x)/H_{\nu },&\nu <M\\(g_{M}(x)-g_{M}^{*})/H_{M},&\nu
=M\end{cases}}
  \end{array}
\end{displaymath}

Here \(M=\max_{}^{}\left\{\nu (x):x\in [0,1]\right\}\) and \(g_{M}^{*}=\min
_{}^{}\left\{g_{M}(x):x\in \cap _{i=0}^{M-1}Q_{i}\right\}\).
Because of the definition of the number \(M\), the problem of finding \(g_{M}^{*}\) always
has a solution. If \(M=m+1\), then \(g_{M}^{*}=\varphi(x^{*})\).
The arcs of the functions \(\psi (x)\) are the H\"{o}lder ones within the sets
\(\cap _{i=0}^{j}Q_{i},0\leq j\leq M-1\)
with the constant 1, and \(\psi (x)\) itself can have the breaks of the first kind at the boundaries
of these sets.
Despite the values of the H\"{o}lder constants \(H_k\) and the value \(g_{M}^{*}\) are not
known in advance, these ones can be estimated in the course of solving the problem.

@@

At the first iteration of the method, the trial is performed in the arbitrary internal point \(x_1\) of
the interval \([0,1]\).
The indices of the points 0 and 1 are considered to be zero ones, the values \(z\) in these points
are undefined.
Let \(k\geqslant 1\) iterations of the method have been performed, in the course of which the
trials were conducted in \(k\) points \(x_i, 1\leqslant i\leqslant k\).
Then, the points \(x^{k+1}\) of the search trials of the next \((k+1)^{\rm th}\) iteration are
defined according to the rules:

Step 1. Renumber the points of the set \(X_k=\{x^1,\dotsc,x^k\}\cup\{0\}\cup\{1\}\), which
includes the boundary points of the interval \([0,1]\) as well as the points of preceding trials by
the lower indices in the order of increasing coordinate values i. e.
\begin{equation}
  \label{eq:points}
  0=x_0<x_1<\dotsc<x_{k+1}=1
\end{equation}
and juxtapose the values \(z_{i}=g_{\nu }(x_{i}),\nu =\nu (x_{i}),i={\overline {1,k}}\) to these
ones.

@@

Step 7. Perform the next trial in the middle of the interval \((x_{t-1};x_{t})\) if the indices of its
end points are not the same: \(x^{k+1}={\frac {1}{2}}(x_{t}+x_{t-1})\).
Otherwise, perform the trial at the point
\begin{displaymath}
  x^{k+1}={\frac {1}{2}}(x_{t}+x_{t-1})-\operatorname {sgn}(z_{t}-z_{t-1}){\frac {|z_{t}-
z_{t-1}|^{n}}{2r_{\nu }\mu _{\nu }^{n}}},\nu =\nu (x_{t})=\nu (x_{t-1}),
\end{displaymath}
and then increment \(k\) by 1.
